---
title: "Export to Gephi"
author: "Justin Bogias"
date: "`r format(Sys.Date(), '%d %b, %Y')`"
output: 
  rmdformats::readthedown
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Setup
```{r}
library(rmdformats)
library(rmarkdown)
library(dplyr)
library(readr)
library(tibble)
library(magrittr)
library(WGCNA)
library(here)

placenta_data <- read_csv(here("data/placenta_data_wgcna.csv")) %>%
  column_to_rownames("sample_id")

gene_entrez_anno <- read_csv(here("data/gene_extrez_anno.csv.gz"))

mergedColours <- read_rds(here("data/mergedColours.rds"))

soft_power <- 12
```

# Export to Gephi
Finally, we want to visualize our networks so we are exporting them as node and edge files that we can use in Cytoscape/gephi or other visualizing tools. 
## Adjacency Matrix
First form the adjacency matrix
```{r export_gephi, eval=TRUE}
# Changed adjacency threshold from 0.5
adj_mat <- adjacency(placenta_data,
                     power = soft_power,
                     type = "signed")
```

Next we will export the data in a Cytoscape format, as this will also allow us to visualise the network in Gephi
```{r eval=FALSE}
exportNetworkToCytoscape(
  adjMat = adj_mat,
  edgeFile = here("data/GDM_EdgeFile_lowfilter.tsv"),
  nodeFile = here("data/GDM_NodeFile_lowfilter.tsv"),
  weighted = TRUE,
  threshold = 0.2,
  nodeNames = NULL,
  altNodeNames = NULL,
  nodeAttr = mergedColours,
  includeColNames = TRUE)
```

## Nodes
Next we want to generate the nodes of our network
```{r}
nodesPlusAnnot <- read_delim(
  here("data/GDM_NodeFile_lowfilter.tsv"),
  delim = "\t") %>%
  dplyr::select(-altName) %>%
  set_colnames(c("id", "colour")) %>%
  left_join(gene_entrez_anno %>%
              dplyr::select("id" = "gene_id",
                            "gene" = "gene_name"),
            by = "id") %>%
  dplyr::filter(!colour == "grey") %>%
  write_csv(here("data/GDM_NodeFile_lowfilter.csv"))
```

## Edges
```{r}
edgePlusAnnot <- read_delim(
  here("data/GDM_EdgeFile_lowfilter.tsv"),
  delim = "\t") %>%
  left_join(gene_entrez_anno %>%
              dplyr::select("fromNode" = "gene_id",
                            "SourceAltName" = "gene_name"),
            by = "fromNode") %>% 
  dplyr::select("Source" = "fromNode",
                "Target" = "toNode",
                "weight",
                "direction",
                "SourceAltName") %>%
  left_join(
    gene_entrez_anno %>%
      dplyr::select("Target" = "gene_id",
                    "TargetAltName" = "gene_name")
  ) %>%
  dplyr::filter(
    Source %in% nodesPlusAnnot$id & Target %in% nodesPlusAnnot$id
  ) %>%
  write_csv(here("data/GDM_EdgeFile_lowfilter.csv"))

```

## Green and Pink module retention in network
How many green and pink were retained in the network
```{r network_nodes_edges, eval=TRUE}
pink_green_nodes <- nodesPlusAnnot %>%
  set_colnames(c("nodeName", "altName", "module")) %>% 
  dplyr::filter(module %in% c("pink", "green"))

edge_from_node <- edgePlusAnnot %>%
  dplyr::filter(Source %in% pink_green_nodes$nodeName |
                  Target %in% pink_green_nodes$nodeName) %>%
  dplyr::filter(weight >= 0.4) %>%
  dplyr::count(Source) %>%
  dplyr::filter(n > 3) %>%
  left_join(pink_green_nodes,
            by = c("Source" = "nodeName"))

edge_to_node <- edgePlusAnnot %>%
  dplyr::filter(Source %in% pink_green_nodes$nodeName |
                  Target %in% pink_green_nodes$nodeName) %>%
  dplyr::filter(weight >= 0.4) %>%
  dplyr::count(Target) %>%
  dplyr::filter(n > 3) %>%
  left_join(pink_green_nodes,
            by = c("Target" = "nodeName"))

gene_node_modules <- edge_from_node %>%
  full_join(edge_to_node,
            by = c("Source" = "Target")) %>%
  dplyr::select("Gene" = "Source",
                "module.x",
                "module.y")

coalesce(gene_node_modules$module.x,
         gene_node_modules$module.y) %>%
  as.data.frame() %>%
  set_colnames("modules") %>%
  dplyr::count(modules)
```

## The Split Turquoise Group
In the visualised module there were two clusters of turquoise genes.
It would be interesting to know which genes were in each and if they 
represent different biological functions.
There wasn't any way to select them from Gephi, so I just had to write down the
gene names directly from it.

```{r}
left_group <- c("IL2RB", "HN1", "CRABP2", "RASGRF2", "TAC3",
                "GLIPR1", "PLA2G7", "PIPOX", "SERPINE2", "IL1R2",
                "REPS2", "KDELR3", "XCR1", "PTK2B", "EXTL3", "EGLN3",
                "CLDN19", "FAT2", "ASCL2", "NOG", "PRG2", "SFRP1", "NOTUM",
                "FN1", "PCSK6", "ASAP3")
write_rds(left_group, here("data/turquoise_left_group.rds"))

right_group <- c("HEMGN", "EPB42", "EPB41", "SLC25A37", "SLC4A1", "FAM46C",
                 "SNCA", "DMTN", "HBA1", "ALAS2")
write_rds(right_group, here("data/turquoise_right_group.rds"))
```